Mental simulation as Bayesian Quadrature

### 2013-05-31 man/nips-2013.tex
Show all rotations, not means
First pass through results

\paragraph{\Naive{} model}

@@

\paragraph{Parametric (Von Mises) model}

@@

\paragraph{Nonparametric (Bayesian Quadrature) model}

@@

We evaluated each model's performance on 20 randomly generated shapes
which had between 3 and 6 vertices, inclusive, (e.g. $X_a$ in Figure
\ref{fig:stimuli}). For each shape, we computed 18 ``different'' and
18 ``same'' stimuli pairs, with $R$ spaced at $20^\prime$ increments
between 0 and 360. ``Different'' pairs were created by reflecting
$X_a$ across the $y$-axis and rotating it by $R$; for ``same'' pairs,
$X_a$ was rotated by $R$.

We considered three metrics of performance in particular:
\begin{itemize}
\item \textit{Rotations}: for ``same'' pairs, how correlated are the
  true angles of rotation with the rotations performed by the model?
  This is defined to be the Pearson's correlation coefficient $\rho$
  between $R$ and $\vert \mathbf{R}\vert$. Figure \ref{fig:rotations}
  shows individual points corresponding to true rotation vs. rotation
  by the model, for each model.
\item \textit{Responses}: how accurate is the model at choosing the
  correct hypothesis? This is defined to be the mean error ($\ME{}$),
  or fraction of times the model picks the incorrect hypothesis.
\item \textit{Estimates of $Z$}: how accurate is the model's estimate
  of $Z$? This is defined to be the mean squared error ($\MSE{}$)
  between the model's estimate of $Z$ and the ``gold standard'' value,
  where the error has scale such that $\MSE{}=0$ indicates no error,
  and $\MSE{}=1$ indicates maximum error. Figure \ref{fig:accuracy}
  shows plots of the true (gold standard) value of $Z$ vs. the model's
  estimate, for each model.
\end{itemize}

@@

The gold standard always evaluates $S$ at all angles, so the
correlation between it and the true angle of rotation is
undefined. The mean error of responses was $\GSME{}$, indicating that
$Z$ is a sufficient approximation of $\int_R p(X_b\ \vert\ X_R)p(R)\
\mathrm{d}R$.

\paragraph{\Naive{} model}

The correlation between the number of rotations computed by the
\naive{} model and the true angle of rotation was $\Naivecorr{}$. The
shape of the data is more interesting (see Figure \ref{fig:rotations},
left panel): the \naive{} model actually corresponds quite well to the
true angle of rotation for $R<\frac{\pi}{2}$ ($\Naivecorrsm{}$). This
is unsurprising, because the closer the true angle is to zero, the
less the model has to rotate, and the less likely it will get stuck on
local maxima. Thus, it is more likely to locate the global maximum,
which generally corresponds to the true angle of rotation. For
$R>\frac{\pi}{2}$, we see an increasing tendency to under-rotate
($\Naivecorrbg{}$); this is likely because it finds a local maxima and
ends prematurely.

The \naive{} model's response error rate was better than chance, but
still quite high, with $\NaiveME{}$. Closer inspection reveals that
the bulk of this comes from $h_1$ stimuli ($\NaiveMEsame{}$
vs. $\NaiveMEdiff{}$ for $h_0$ stimuli). This is, as above, probably
related to finding only local maxima: if the model finds a local
maxima which is low enough, the area under the estimated curve will
not be large enough to accept $h_1$.

This is reflected in the accuracy in estimating $Z$ as well
($\NaiveMSE{}$). Figure \ref{fig:accuracy} (left panel) shows
individual estimates for each stimulus, color-coded by true
hypothesis. The model underestimates $Z$ for significant portion of
the $h_1$ stimuli. On the other $h_1$ stimuli, the \naive{} model
actually tends to overestimate because linear interpolation does not
account for symmetry of global maxima (e.g., Figure \ref{fig:li}).

\paragraph{Parametric (Von Mises) model}

Like the \naive{} model, the parametric model uses hill-climbing as a
sampling strategy, so they have identical correlations with the true
angle of rotation. However, given the same $\mathbf{R}$, the
parametric model estimates $Z$ differently, thus giving rise to
different response and estimated $Z$ error rates. The response error
for the parametric model was higher than the \naive{} model, with
$\VMME{}$. Breaking this down, we see that the error for $h_1$ stimuli
is $\VMMEsame{}$: in other words, the parametric model is at chance
when determining whether two identical shapes are the same. This is
because it never overestimates (see Figure \ref{fig:accuracy}, center
panel). So, it accurately estimates $Z$ for stimuli which the \naive{}
model would overestimate ($\VMMSE{}$), but still has poor performance
when $Z$ is underestimated. \TODO{Indeed, if we exclude stimuli for
  which the \naive{} model overestimates ($Z\geq 0.3$), the error for
  the \naive{} model lowers to match that of the parametric model.}

\paragraph{Nonparametric (Bayesian Quadrature) model}

The rotations computed by the nonparametric model were strongly
correlated with the true rotations ($\BQcorr$, see Figure
\ref{fig:rotations}, right panel). This is because the nonparametric
model does not get stuck as easily on local optima: it will continue
rotating until it is confident that its estimate of $Z$ is accurate.

Correspondingly, the nonparametric model is much more accurate in
choosing the correct hypothesis ($\BQME{}$). As with the \naive{} and
parametric models, it mostly errs on $h_1$ stimuli ($\BQMEsame$). This
appears to mostly be the result of choosing the incorrect initial
direction of rotation.

Similarly, because the nonparametric model actively attempts to obtain
an accurate estimate of $Z$, its estimate of $Z$ is indeed fairly
accurate ($\BQMSE{}$). This is slightly higher than the nonparametric
model, but is due to a few outliers. \TODO{If we again exclude stimuli
  for which $Z$ is greatly overestimated ($Z\geq 0.3$), the
  nonparametric model's error decreases significantly
  ($\MSE{}=0.03$).}

>>>>

\paragraph{\Naive{}}

@@

\paragraph{Parametric (Von Mises)}

@@

\paragraph{Nonparametric (Bayesian Quadrature)}

@@

We evaluated each model's performance on 20 randomly generated shapes
which had between 3 and 6 vertices, inclusive, (e.g. $X_a$ in Figure
\ref{fig:stimuli}). For each shape, we computed 18 ``different'' and
18 ``same'' stimuli pairs, with $R$ spaced at $20^\prime$ increments
between 0 and 360. ``Different'' pairs were created by reflecting
$X_a$ across the $y$-axis and rotating it by $R$; for ``same'' pairs,
$X_a$ was rotated by $R$. \TODO{need to mention that shapes were
  chosen to be discriminable by the gold standard}

We considered three metrics of performance in particular:
\begin{itemize}
\item \textit{Responses}: how accurate is the model at choosing the
  correct hypothesis? This is defined to be the mean error ($\ME{}$),
  or fraction of times the model picks the incorrect hypothesis.
\item \textit{Rotations}: for those ``same'' pairs which the model
  gets correct, how correlated are the true angles of rotation with
  the rotations performed by the model?  This is defined to be the
  Pearson's correlation coefficient $\rho$ between $R$ and $\vert
  \mathbf{R}\vert$. Figure \ref{fig:rotations} shows individual points
  corresponding to true rotation vs. rotation by the model, for each
  model.
\item \textit{Estimates of $Z$}: how accurate is the model's estimate
  of $Z$? This is defined to be the mean squared error ($\MSE{}$)
  between the model's estimate of $Z$ and the ``gold standard'' value,
  where the error has scale such that $\MSE{}=0$ indicates no error,
  and $\MSE{}=1$ indicates maximum error. Figure \ref{fig:accuracy}
  shows plots of the true (gold standard) value of $Z$ vs. the model's
  estimate, for each model.
\end{itemize}

@@

The mean error of the gold standard model was $\GSME{}$, indicating
that $Z$ is a sufficient approximation of $\int_R p(X_b\ \vert\
X_R)p(R)\ \mathrm{d}R$ for these particular stimuli.

\paragraph{\Naive{}}

The \naive{} model's response error rate was better than chance with
$\NaiveME{}$. Closer inspection reveals that much of this comes from
$h_0$ stimuli ($\NaiveMEdiff{}$), where the asymmetric linear
interpolation gives an overestimate of $Z$ (e.g., in Figure
\ref{fig:li}). There is some error from $h_1$ stimuli as well
($\NaiveMEsame{}$), which happens when the hill climbing search finds
a local maxima and terminates prematurely, thus resulting in a low
estimate of $Z$.

The correlation between the average number of rotations computed by
the \naive{} model and the true angle of rotation was
$\Naivecorr{}$. The shape of the data is more interesting (see Figure
\ref{fig:rotations}, top left panel): the \naive{} model actually
corresponds extremeley well to the true angle of rotation for
$R<\frac{\pi}{2}$ ($\Naivecorrsm{}$). This is unsurprising, because
the closer the true angle is to zero, the less the model has to
rotate, and the less likely it will get stuck on local maxima. Thus,
it is more likely to locate the global maximum, which corresponds to
the true angle of rotation. For $R>\frac{\pi}{2}$, we see an
increasing tendency to under-rotate due to getting stuck on local
maxima, as well as a tendency to over-rotate if the wrong rotation
direction was chosen.

The \naive{} model's error on $h_0$ stimuli is reflected in its
accuracy in estimating $Z$ ($\NaiveMSE{}$). Figure \ref{fig:accuracy}
(left panel) shows individual estimates for each stimulus, color-coded
by true hypothesis. The model overestimates $Z$ for most of the $h_0$
and $h_1$ stimuli, and underestimates several of the $h_1$ stimuli as
well.

\paragraph{Parametric (Von Mises)}

The parametric model is also fairly inaccurate ($\VMME{}$), making
more errors on $h_1$ stimuli ($\VMMEsame{}$) than $h_0$ stimuli
($\VMMEdiff{}$). This is because the Von Mises distribution only has a
single peak and thus will underestimate $Z$ for multimodal similarity
functions. However, it does more accurately estimate $Z$ for stimuli
that the \naive{} model would overestimate ($\VMMSE{}$). This
underestimation is severe enough that in most cases where it gets
stuck on a local maxima, it produces the incorrect judgment. So, the
correlation for \textit{correct} responses is actually improved
($\VMcorr{}$) and would be near perfect if the correct diretion of
rotation was initially chosen (see Figure \ref{fig:rotations}, top
center).

\paragraph{Nonparametric (Bayesian Quadrature)}

The nonparametric model is much more accurate in choosing the correct
hypothesis ($\BQME{}$), erring approximately equally for both types of
stimuli. We note that this is very close to the 3.2\% error rate
reported by \cite{Shepard1971}, though further experimentation is
necessary to determine whether the types of errors people make align
with those of the model.

The average rotations computed by the nonparametric model were
strongly correlated with the true rotations ($\BQcorr$, see Figure
\ref{fig:rotations}, bottom right). This is because the nonparametric
model does not get stuck as easily on local optima: it will continue
rotating until it is confident that its estimate of $Z$ is
accurate. Indeed, its active attempts to obtain an accurate estimate
of $Z$ are largely succeessful ($\BQMSE{}$). Moreover, it has the
capacity to ``restart'' and can recover from initially rotating in the
incorrect direction, as illustrated by the fewer number of extreme
over-rotations in Figure \ref{fig:rotations} (top right).
